RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-03 manuscripts/manuscript.Rmd
Rewrite of first part of the simmap section.

Here we illustrate how NeXML's metadata system, and `RNeXML`'s API to
it, can be used to record and share a type of phylogenetic data not
taken into account when NeXML was designed. Specifically, we use the
example of stochastic character maps [@Huelsenbeck_2003], which assign
sections of each branch in a phylogeny to a particular "state"
(typically of a morphological characteristic). The current de-facto
format for sharing stochastic character maps is a (non-interoperable)
modification of the standard Newick tree format created by a software
tool widely used for creating such maps, `simmap` [@Bollback_2006].

@Revell_2012 describes the `phytools` package for R, which includes
utilities for reading, manipulating, and writing `simmap` files in R.
In this example, we also show how to define `RNeXML` functions that 
map the R representation used by Revell (an extension of the `ape` class)
into the NeXML extension we have defined by using `RNeXML` functions.

Since a stochastic character map simply assigns different states to
parts of a branch (or edge) on the phylogenetic tree, we can create
a NeXML representation by annotating the `edge` elements with appropriate
`meta` elements.  These elements need to describe the character state
being assigned and the duration (in terms of branch-length) that the edge 
spends in that state (Stochastic character maps are specific to time-calibrated
or ultrametric trees).

NeXML already defines the `characters` element to handle discrete character traits (`nex:char`)
and the states they can assume (`nex:state`).  We will thus reuse the `characters` element for
this purpose, referring to both the character trait and the states by the ids assigned to them
in that element.  (NeXML's convention of referring to everything by id permits a single canonical
definition of each term, making it clear where additional annotation belongs).  For each edge, we 
need to indicate:

By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

<!-- Ugh, do we need to explain that everything in NeXML uses an id so that it can be reused in this way?
     -->

- That our annotation contains a stochastic character mapping reconstruction
- Since many reconstructions are possible for a single edge, we give each reconstruction an id
- We indicate for which character trait we are defining the reconstruction 
- We then indicate which states the character assumes on that edge. 
  For each state realized on the edge, that involves stating: 
    + the state assignment
    + the duration (length of time) for which the edge spends in the given state
    + the order in which the state changes happen (Though we could just assume 
      state transitions are listed chronologically, NeXML suggests making all 
      data explicit, rather than relying on the structure of the data file to
      convey information).

Thus the annotation for an edge that switches from state `s2` to state 
`s1` of character `cr1` would be constructed like this:

>>>>

Here we illustrate how NeXML's metadata system, and `RNeXML`'s API to
it, can be used to record and share a type of phylogenetic data not
taken into account when NeXML was designed. Specifically, we use the
example of stochastic character maps [@Huelsenbeck_2003], which assign
certain parts (corresponding to time) of each branch in a
time-calibrated phylogeny to a particular "state" (typically of a
morphological characteristic). The current de-facto format for sharing
stochastic character maps is a (non-interoperable) modification of the
standard Newick tree format created by a software tool widely used for
creating such maps, `simmap` [@Bollback_2006].

To express stochastic character map data as a custom metadata
extension, we create metadata annotations for the `edge` elements in
the phylogeny topology that relate parts of the annotated edge (branch)
to different states. These annotations need to describe the character state
being assigned, and the duration (in terms of branch length) that the edge 
spends in that state. The `NeXML` standard already defines `state`
elements for discrete character traits, which, as most elements in
NeXML, also have unique identifies (attribute `id`) by which they can
be referenced in other parts of the document. In the following
proposal, we call each series of character state assignments for an edge a
_reconstruction_, and for each state assignment comprising the
reconstruction, we provide the length, the identifier of the character
state, and an ordinal property to record the chronological order of
the state assignment (the identifier values in the proposal, "cr1",
"s1", and "s2", are -- fictitious -- identifiers that, in contrast to
global identifiers, are unique only within a given NeXML document):
